On Devon Allen's Disqualification at the 2022 World Track and Field Championships

### 2023-09-06 Manuscript/manuscript.tex
One comment
Started to change language and add 2023 data

Devon Allen, an alumnus of the University of Oregon in 
Eugene, Oregon, was
expected to be a contender for the 2022 World Track and Field Championships 110 
meter hurdle.  After running the event in 12.84 seconds in early June 2022 
(only 0.04 seconds away from the world record \citep{wa2022preview}) and coming 
in third at the United States Track and Field 
Championships also in Eugene in late June, Allen was poised to compete in the 
2022 World Track and Field Championships at the same Eugene venue in August.
At the World Championships, Allen made it through the preliminary
heats and semifinals to reach the final heat while posting reaction times of 
0.123 and 0.101 in his preliminary heat and semifinal, respectively.  
His disqualification in the final heat by reacting in 0.099 seconds, just 0.001
seconds faster than the 0.1 threshold set by the International Association of
Athletics Federations (IAAF), was met
with shock from both Allen and fans alike.  NBC, who was broadcasting the
championships in the United States, later uploaded a video to their YouTube
channel that showed the disqualification
(\url{https://www.youtube.com/watch?v=D6NXTMo-1yM}).
The crowd erupted with jeers
when they found out that Allen was disqualified, angry at the result.  Had Allen 
reacted just 0.001 seconds slower he would have been allowed to compete and been 
spared waiting until the next World Track and Field Championships.
\eds{He is not competing this year, correct? If yes, we may need to rephrase.}

@@

There are two types of data that are explored throughout this paper: data from
1999-2022 for every world championship, and data from athletes who competed in
the 2022 World Championships and another competition (either 2019 World 
Championships or a 2022 national competition typically held within three months 
of the 2022 World Championships). 
%\eds{is this true? are you still using the 2019 world championships data for the
%rank-based analyses, or only the national competitions?}
%\of{Yes the ranked based comparison has 2 subsections: 2022 national vs 2022
%international and 2019 international vs 2022 international.  Should I instead say
%that there are three types of data across 2 sections or something like that?}
%\eds{I think this is fine here, but it is not at all clear that there are two
%subsections in the rank-based results for 2022 national vs 2022
%international and 2019 international vs 2022 international.}
%\of{Currently under the Methods/Results section I have GLMM and rank based comparison
%do you think I need subsections under rank based comparison, or is that too much
%compartmentalization?  Or should I instead say there are 3 types of data that are
%used in two different methods, or something like that?} \eds{I prefer
%subsections within rank-based comparison.}
The first data set is used for the GLMM analyses described in 
Section~\ref{sec:glmm}, while the second collection of data is used for the 
rank-based comparisons described in Section~\ref{sec:rank}.

@@

For analysis purposes, we will often pool reaction times of men's final and 
men's semifinal heats together which increases the sample size. This is 
advantageous because in some years there are very few finals observations, such
as in 2022, when there were only five data points due to two disqualifications 
and one
athlete who did not compete.  We also consider data that
excludes 2022 to see how our analysis differs based on the exclusion of one
particular year of interest.

The data under consideration is summarized using a boxplots in 
Figure~\ref{fig:Boxplot}.  It is clear that the reaction times in the 
2022 boxplot for both 
the men's finals and men's pooled semifinal and final data are lower (indicating 
faster reaction times)
than in previous years.  The median for the men's pooled data in 2022 was 0.129
and the median for the men's final data in 2022 was 0.126.  For comparison,
\cite{brosnan2017effects} found median reaction times of 0.156 seconds 
for men when looking at data from 1999 to 2014.
We can also see from Figure~\ref{fig:Boxplot} that the reaction times clearly
differ from year to year. The venue of the World Championships changes each 
year, so weather or climate related factors such as humidity, precipitation, 
elevation, may indeed play a role, but additionally, both the technology and 
false start penalization rules have changed over the study period.

\begin{figure}[tbp]
  \centering
  \includegraphics{Boxplot}
  \caption{The reaction times from 1999 to 2022 for the men's final (top) and
  the men's semifinals and finals (bottom). 
	\eds{there is only one plot here, not two. If you are keeping only one, the
	text and caption need to be modified accordingly}}
  \label{fig:Boxplot}
\end{figure}

@@

\subsubsection{2022 National Competitions}\label{sec:datanational}
From June 23-26, 2022, the United States held its Track and Field Championships 
at Hayward Field in Eugene, Oregon to decide who to send to the World 
Championships being held at the same venue in August. Thus, we can establish a 
baseline for the four United States athletes who competed in at least one round 
of both events: Trey Cunningham, Daniel Roberts, Grant Holloway, and Devon Allen.   
Every athlete reacted faster in all the World Track and Field Championships 
races compared to the United States Track and Field Championships races. The 
dataset of the four United States athletes is too small to perform any 
meaningful statistical analysis, but the data was expanded to include reaction 
times for 11 more athletes who competed at other national competitions (many of which were 
held from May-July of 2022). This data was difficult to find, as the data is not
centrally located but instead found on different national websites (often 
recorded in the national language without translation, and hence not used). 
\of{I don't know how to say this}. \eds{is my change correct?} Each athlete was 
designated as a ``cluster'' and the stage of competition (national or international) 
was specified for each competition within the cluster (athlete).
Thus, this is subunit level grouping where we assign the treatment, national or
international, within each cluster.  Cluster sizes range from three to six, with
one cluster of size three, six clusters of size four, two clusters of size five,
and two clusters of size six.  We chose to include preliminary rounds in this
section as well as in Section~\ref{sec:data2019} in order to increase the size 
of our data.  Without data from the preliminary rounds, the number of observations
dropped from 65 to 26 and the number of athletes included in the analysis
decreased from 15 to 9.

@@

\subsubsection{2019 World Track and Field Championships}\label{sec:data2019}
The 2019 World Track and Field Championship data is included in the larger dataset
described in Section~\ref{sec:dataworld}, but we can also use it similarly to
how we used national level in Section~\ref{sec:datanational}.  We can compare
the times of athletes who participated in the 2019 and 2022 World Track and Field
Championships to see if there were lower times in 2022.  While it is theoretically
possible that athletes got faster between 2019 and 2022, it is important to
remember that these are some of the fastest sprinters on the planet, and thus
improving on elite times is not easy.  Once again we repeat a similar procedure
as in Section~\ref{sec:datanational}; we assign each athlete to be a cluster and
within each cluster label times to be from either the 2019 World Championships
or the 2022 World Championships.  In this analysis, we look at data from 15
athletes, where cluster size ranges from two to six.   There were two clusters 
of size two, three clusters of size three, five clusters of size four, two 
clusters of size five, and three clusters of size six.

\jy{give specifics of your sample size}

\begin{figure}[tbp]
  \centering
  \includegraphics{Scatterplot2019vs2022}
  \caption{The reaction times from athletes who competed at least once in both 
  the 2019 and 2022 World Track and Field Championships}
  \label{fig:Scatterplot2019vs2022}
\end{figure}
\of{I included this here for now, but feel free to change the location.  I will
add references to it and explanation of it along with some descriptive statistics about
the size and mean/median,std dev, size of the data.}

Figure~\ref{fig:Scatterplot2019vs2022} shows the distribution of times from the
2019 and 2022 World Championships.  
\eds{Please reorganize the labels so they are either in one of the following 
orders:
2019 Heat, 2019 Semi, 2019 final, 2022 Heat, 2022 Semi, 2022 final  OR
2019 Heat, 2022 Heat, 2019 Semi, 2022 Semi, 2019 final, 2022 final.  
(I am not sure which is better, but perhaps the latter?)}
This data was only for athletes who competed
in at least one race at each championship, however it is possible for athletes
to have competed in one race in 2019 and two races in 2022.  The 2022 World
Championship Heats and Semifinals both had on average lower reaction times than
their respective 2019 races, and the 2022 finals were substantially faster. Also
worth noting is that the fastest reaction time in the 2022 Finals and 2022
Semifinals both came from Devon Allen, but the difference between 0.101 and 0.099
seconds was the difference in what caused him to be disqualified. 
%\eds{check these numbers? 0.099?}\of{resolved}

@@

A summary of the two effect model is shown in Table~\ref{tab:Gamma_parameters}
for the pooled datasets and shows the standard error of the year
effect increases for both the men's final and men's pooled data sets when
including 2022. We choose to only look at the pooled datasets, as R displayed
warnings when creating the finals only models.  The two effect model with venue
and heat effect was the best model, as indicated by an ANOVA test 
\eds{I think it may be better to call this a likelihood ratio test rather than
an ANOVA test, assuming that is the formal test being used} and decrease in
AIC.  For the entire pooled dataset, the AIC for the venue effect only model was
$-1847.9$, the AIC for the heat effect only model $-1894.4$, and the AIC for the 
two effect model, which we will present throughout this section, was $-1898.9$. 
Thus, the two
effect model is superior and will be presented throughout the rest of the paper.
\of{How specific do I need to be in this section? Will people who read this paper
run the code to see the p value or should I say it directly and how do I say it?}
\eds{If you are doing a likelihood ratio test, I suggest including the p-value 
and the degrees of freedom associated with the test.}

\begin{table}
  \centering
  \caption{Summary statistics for the two effect GLMM on various data sets. SE
  is short for standard error. \eds{data in table needs to be revisited and table
	needs to be referenced within the text. Also, it needs to be clarified 
	somewhere that ``Old Pooled'' excludes 2022 and ``All Pooled'' includes 2022.}} 
  \begin{tabular}{c c c c c c}
      \toprule
      Dataset & AIC & $\sigma_h^2$ & $\sigma_v^2$ & $\alpha$ & $\delta_{i/j}$ \\
      \midrule
      Old Pooled & 0.060 & 0.034 & $-1.865$ & 0.139 \\
      All Pooled & 0.059 & 0.041 & $-1.880$ & 0.014 \\
      \bottomrule
  \end{tabular}
  \label{tab:Gamma_parameters}
\end{table}

>>>>

Devon Allen, an alumnus of the University of Oregon in 
Eugene, Oregon, was
expected to be a contender for the 2022 World Track and Field Championships 110 
meter hurdle.  After running the event in 12.84 seconds in early June 2022 
(only 0.04 seconds away from the world record \citep{wa2022preview}) and coming 
in third at the United States Track and Field 
Championships also in Eugene in late June, Allen was poised to compete in the 
2022 World Track and Field Championships at the same Eugene venue in August.
At the World Championships, Allen made it through the preliminary
heats and semifinals to reach the final heat while posting reaction times of 
0.123 and 0.101 in his preliminary heat and semifinal, respectively.  
His disqualification in the final heat by reacting in 0.099 seconds, just 0.001
seconds faster than the 0.1 threshold set by the International Association of
Athletics Federations (IAAF), was met
with shock from both Allen and fans alike.  NBC, who was broadcasting the
championships in the United States, later uploaded a video to their YouTube
channel that showed the disqualification
(\url{https://www.youtube.com/watch?v=D6NXTMo-1yM}).
The crowd erupted with jeers
when they found out that Allen was disqualified, angry at the result.  Had Allen 
reacted just 0.001 seconds slower he would have been allowed to compete and been 
spared waiting until the next World Track and Field Championships.
\eds{He is not competing this year, correct? If yes, we may need to rephrase.}
\of{Correct}

@@

There are two types of data that are explored throughout this paper: data from
1999-2023 for every world championship, and data from athletes who competed in
the 2022 World Championships and another competition (either 2019 World 
Championships, 2023 World Championships or a 2022 national competition typically
held within three months of the 2022 World Championships). 
%\eds{is this true? are you still using the 2019 world championships data for the
%rank-based analyses, or only the national competitions?}
%\of{Yes the ranked based comparison has 2 subsections: 2022 national vs 2022
%international and 2019 international vs 2022 international.  Should I instead say
%that there are three types of data across 2 sections or something like that?}
%\eds{I think this is fine here, but it is not at all clear that there are two
%subsections in the rank-based results for 2022 national vs 2022
%international and 2019 international vs 2022 international.}
%\of{Currently under the Methods/Results section I have GLMM and rank based comparison
%do you think I need subsections under rank based comparison, or is that too much
%compartmentalization?  Or should I instead say there are 3 types of data that are
%used in two different methods, or something like that?} \eds{I prefer
%subsections within rank-based comparison.}
The first data set is used for the GLMM analyses described in 
Section~\ref{sec:glmm}, while the second collection of data is used for the 
rank-based comparisons described in Section~\ref{sec:rank}.  This paper was
started before the 2023 World Championships, but as that data has become
available recently, we are pleased to include it in our analysis.

@@

For analysis purposes, we will often pool reaction times of men's final and 
men's semifinal heats together which increases the sample size. This is 
advantageous because in some years there are very few finals observations, such
as in 2022, when there were only five data points due to two disqualifications 
and one athlete who did not compete.  This is the data that we will reference
throughout the rest of the paper unless otherwise noted. We also consider data 
that excludes 2022 to see how our analysis differs based on the exclusion of one
particular year of interest.

The data under consideration is summarized using a box plot in 
Figure~\ref{fig:Boxplot}.  It is clear that the reaction times in the 2022
box plot are lower (indicating faster reaction times) than in many of the other
years.  The median reaction time in 2022 was 0.129 seconds,  compared to 0.156
seconds, which is what \citet{brosnan2017effects} found when looking at data
from 1999 to 2014. We can also see from Figure~\ref{fig:Boxplot} that the 
reaction times clearly differ from year to year. The venue of the World 
Championships changes each year, so weather or climate related factors such as 
humidity, precipitation, elevation, may indeed play a role, but additionally,
both the technology and false start penalization rules have changed over the 
study period.

\begin{figure}[tbp]
  \centering
  \includegraphics{Boxplot}
  \caption{The reaction times from 1999 to 2023 for the men's 110 meter hurdle.
	\eds{there is only one plot here, not two. If you are keeping only one, the
	text and caption need to be modified accordingly}}
  \of{Resolved}
  \label{fig:Boxplot}
\end{figure}

@@

\subsubsection{2022 National Competitions}\label{sec:datanational}
From June 23-26, 2022, the United States held its Track and Field Championships 
at Hayward Field in Eugene, Oregon to decide who to send to the World 
Championships being held at the same venue in August. Thus, we can establish a 
baseline for the four United States athletes who competed in at least one round 
of both events: Trey Cunningham, Daniel Roberts, Grant Holloway, and Devon Allen.   
Every athlete reacted faster in all the World Track and Field Championships 
races compared to the United States Track and Field Championships races. The 
dataset of the four United States athletes is too small to perform any 
meaningful statistical analysis, but the data was expanded to include reaction 
times for 11 more athletes who competed at other national competitions (many of which were 
held from May-July of 2022). This data was difficult to find, as the data is not
centrally located but instead found on different national websites (often 
recorded in the national language without translation, and hence not used). 
\of{I don't know how to say this}. \eds{is my change correct?} \of{No I was able
to use a lot even though I couldn't actually read the page. I just tried to find
110 meter hurdle and then look at a column for reaction times which are always from
roughly 0.1 to 0.18 seconds.} Each athlete was 
designated as a ``cluster'' and the stage of competition (national or international) 
was specified for each competition within the cluster (athlete).
Thus, this is subunit level grouping where we assign the treatment, national or
international, within each cluster.  Cluster sizes range from three to six, with
one cluster of size three, six clusters of size four, two clusters of size five,
and two clusters of size six.  We chose to include preliminary rounds in this
section as well as in Section~\ref{sec:data2019} in order to increase the size 
of our data.  Without data from the preliminary rounds, the number of observations
dropped from 65 to 26 and the number of athletes included in the analysis
decreased from 15 to 9.

@@

\subsubsection{Other World Track and Field Championships}\label{sec:data2019}
The 2019 and 2023 World Track and Field Championships data is included in the larger dataset
described in Section~\ref{sec:dataworld}, but we can also use it similarly to
how we used national level in Section~\ref{sec:datanational}.  We can compare
the times of athletes who participated in the 2022 Championships with their times
at another World Championships: 2019 and/or 2023.

While it is theoretically possible that athletes got faster 
between 2019 and 2022 or 2022 and 2023, it is important to 
remember that these are some of the fastest sprinters on the planet, and thus
improving on elite times is not easy.  Once again we repeat a similar procedure
as in Section~\ref{sec:datanational}; we assign each athlete to be a cluster and
within each cluster label times to be from either the 2019, 2022, or 2023 World
Championships. When comparing 2019 and 2022 we look at data from 15
athletes, with two clusters of size two, three clusters of size three, five 
clusters of size four, two clusters of size five, and three clusters of size six.
When comparing 2022 and 2023 we look at data from 22 athletes, with seven 
clusters of size two, three clusters of size three, six clusters of size four,
and six clusters of size five.

\jy{give specifics of your sample size}
\of{Resolved}

\begin{figure}[tbp]
  \centering
  \includegraphics{RankScatterPlots}
  \caption{The top graph shows the athletes who competed at least once each at
  the 2019 and 2022 World Championships.  The middle graph shows the athletes
  who competed at least once each at the 2022 and 2023 World Championships. The
  bottom graph shows the athletes who competed at least once at their country's
  2022 national championships and the 2022 World Championship.}
  \label{fig:RankScatterplots}
\end{figure}

Figure~\ref{fig:RankScatterplots} shows the data that is being analyzed
with the rank-based methods.  In each of the three graphs, athletes who competed
in the 2022 finals are having their times compared to another event they competed
at.  It is worth noting that each instance of the 2022 data is different, because
those who competed at both the 2019 and 2022 World Championships will be
different athletes than those who competed at both the 2022 and 2023 World
Championships. This data was only for athletes who competed
in at least one race at each championship, however it is possible for athletes
to have competed in one race in 2019 and two races in 2022.  The 2022 World
Championship Heats and Semifinals both had on average lower reaction times than
their respective 2019 races, and the 2022 finals were substantially faster. It
is also worth noting is that the fastest reaction time in the 2022 Finals and 2022
Semifinals both came from Devon Allen, but the difference between 0.101 and 0.099
seconds was the difference in what caused him to be disqualified. 
\eds{Please reorganize the labels so they are either in one of the following 
orders:
2019 Heat, 2019 Semi, 2019 final, 2022 Heat, 2022 Semi, 2022 final  OR
2019 Heat, 2022 Heat, 2019 Semi, 2022 Semi, 2019 final, 2022 final.  
(I am not sure which is better, but perhaps the latter?)}
of{I re-worked this graph completely, please let me know what you think.  For
each graph I currently list them in chronological order: so 2022 then 2023 but
I list 2022 second when comparing it to 2019/rank comparison.  Please let me
know your thoughts on this and if I should list 2022 and then 2023 for visual
consistency.}

%\eds{check these numbers? 0.099?}\of{resolved}

@@

A summary of the two effect model is shown in Table~\ref{tab:Gamma_parameters}
for the pooled datasets and shows the standard error of the year
effect increases for both the men's final and men's pooled data sets when
including 2022. We choose to only look at the pooled datasets, as R displayed
warnings when creating the finals only models.  The two effect model with venue
and heat effect was the best model, as indicated by an ANOVA test 
\eds{I think it may be better to call this a likelihood ratio test rather than
an ANOVA test, assuming that is the formal test being used} and decrease in
AIC.  For the entire pooled dataset, the AIC for the venue effect only model was
$-2023.5$, the AIC for the heat effect only model $-2070.7$, and the AIC for the 
two effect model, which we will present throughout this section, was $-2075.7$. 
Thus, the two
effect model is superior and will be presented throughout the rest of the paper.
\of{How specific do I need to be in this section? Will people who read this paper
run the code to see the p value or should I say it directly and how do I say it?}
\eds{If you are doing a likelihood ratio test, I suggest including the p-value 
and the degrees of freedom associated with the test.}

\begin{table}
  \centering
  \caption{Summary statistics for the two effect GLMM on various data sets. SE
  is short for standard error. \eds{data in table needs to be revisited and table
	needs to be referenced within the text. Also, it needs to be clarified 
	somewhere that ``Old Pooled'' excludes 2022 and ``All Pooled'' includes 2022.}} 
  \begin{tabular}{c c c c c c}
      \toprule
      Dataset & AIC & $\sigma_h^2$ & $\sigma_v^2$ & $\alpha$ & $\delta_{i/j}$ \\
      \midrule
      Excluding 2022 &  0.057 & 0.033 & $-1.868$ & 0.136 \\
      All Data & $-2075.7$ 0.057 & 0.040 & $-1.880$ & 0.013 \\
      \bottomrule
  \end{tabular}
  \label{tab:Gamma_parameters}
\end{table}
